---
title: "EXST 7151 Midterm Project"
author: "Dina Dinh"
date: '2023-10-18'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Libraries
```{r}
library(ggplot2)
library(dplyr)
```

Data Cleaning

```{r}
data <- read.csv("data_date.csv")
unique_data <- data %>% distinct(Country, .keep_all = TRUE)
aqi_2 <- unique_data[,c("Status","AQI_Value")]
aqi <- aqi_2[,2]
aqi <-  aqi[aqi <= 500]
aqi_df <- data.frame(aqi)
```

We see that the variable AQI contains no missing values need the be addressed. There is a value greater than 500 which should be removed that's probably due to typing error since AQI only ranges from 0 to 500.

Histogram of AQI

```{r}
hist_aqi <- ggplot(aqi_df, aes(x = aqi)) + geom_histogram(aes(y = ..count..),
breaks = seq(0, 550, by = 25), colour = "black",
fill = "pink") +
geom_density(aes(y = ..count.. * 25), color = "blue", linewidth = 1) +
scale_x_continuous(breaks = seq(0,
550, by = 50)) + theme_classic()

hist_aqi
```

Since the dataset can only take any positive values ranging from 0 to 500, the Gamma distribution seem the most plausible for this dataset. Also, the dataset was recorded over the span of about one year from July 21, 2022, to August 10, 2023.


```{r}
a = 5

x <- seq(0, 50, by = 0.1) 

curve(dgamma(x, shape = a, rate = a/70), col = "grey", lwd = 3, xlab = "X", ylab = "Density",
      main = "Gamma Distribution", ylim = c(0, .08), xlim = c(0, 300))
curve(dgamma(x, shape = sum(aqi) + 1, rate = 142), add = TRUE, col = "green", lwd = 5)
curve(dgamma(x, shape = sum(aqi) + a, rate = 142 + a/70),add=T,col="red",lwd=3)
legend(200,.08,c("Prior","Likelihood", "Posterior"),lty=1,lwd=4,col=c("grey","green","red"))

```

Summary Statistics of the Posterior Distribution
```{r}
mu_n <- ((sum(aqi))+a)/(142+(a/70))
var_n <- (sum(aqi)+a)/((142+a/70)^2)
ci_5 <- qgamma(0.05, shape = sum(aqi) + a, rate = 142 + a/70)
ci_95 <- qgamma(0.95, shape = sum(aqi) + a, rate = 142 + a/70)

mu <- (sum(aqi)+1)/(142)
mu0 <- (a)/(a/70)
```


Predictive
```{r}
y =  seq(0, 500, by = 1) 
z =dnbinom(y, size = a+sum(aqi), mu = (a+sum(aqi))/((142+(a/70))+142))
df_z <- data.frame(z)
barplot(z, names.arg = y, xlab = "AQI", ylab = "Probability", col = "skyblue", main = "Negative Binomial Distribution")
```


